Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-05-16 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Bench scale ch

@@

Three replicate time to ignition experiments were conducted with the cone calorimeter apparatus at each heat flux.  The results of these experiments are shown in Table~\ref{tab:Gypsum_wallboard_timetoign}.  The time to ignition values in the table are the average of the three times.  Note that no ignition occurred at 25~kg/m$^2$.  The exposure at 25~kg/m$^2$ resulted in charring and cracking of the painted paper surface of the gypsum wallboard sample.  This began to occur, on average, at 270 s of exposure time. The samples were exposed to 25~kg/m$^2$ for 1800~s each and no ignition occurred.  In the experiments with heat fluxes of 35~kg/m$^2$ or higher, ignition did occur and the painted paper surface of the gypsum board was consumed, leaving the gypsum board exposed.  There were no surprises in the data with regard to the heat flux exposures.  The time to ignition was reduced as the incident heat flux was increased.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition &  Fire Effect  \\
		\hline (kg/m$^2$) &  & (s)  &   \\
		\hline 25	& No 	& NA 		& Paint \& paper charred \& cracked \\
		\hline 35 	& Yes 	& 73	 	& Paint \& paper consumed \\
		\hline 50	& Yes 	& 35 	 	& Paint \& paper consumed   \\
		\hline 70	& Yes 	& 21 	 	& Paint \& paper consumed \\
		\hline 75	& Yes 	& 17 		& Paint \& paper consumed \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time to ignition results]{Painted gypsum wallboard time to ignition results from the cone calorimeter}
	\label{tab:Gypsum_wallboard_timetoign}
\end{table}

Additional cone calorimeter experiments were conducted in the University of Canterbury Fire Laboratory.  In these experiments the purpose was to obtain an estimate of the temperature of the painted paper surface at the time of ignition.   Samples of painted gypsum wallboard were shipped from the NIST Fire Research Laboratory in Gaithersburg, MD., USA.  In these experiments an additional incident heat flux level of 30~kg/m$^2$ was added to provide another test between the 25 and 35~kg/m$^2$ levels.  The averaged ignition time and temperature pairs are shown in Table~\ref{tab:Gypsum_wallboard_igntemp}.  The average time to ignition values between the two different calorimeters are within $\pm~50~\%$ and $\pm~10~\%$ at incident heat flux values of 35 and 50~kg/m$^2$, respectively.  The average ignition temperatures from the nine experiments ranged from 335 to 375~$^{\circ}$C.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (kg/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 30 	& Yes 	& 112	& 375 	 \\
		\hline 35	& Yes 	& 50 	& 335 	 \\
		\hline 50	& Yes 	& 38 	& 371 	 \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard ignition temperature results]{Painted gypsum wallboard ignition temperature results from exposure in the cone calorimeter}
	\label{tab:Gypsum_wallboard_igntemp}
\end{table}

In the radiant flooring panel experiments the gypsum wallboard was exposed to an incident heat flux that ranged from approximately 21~kg/m$^2$ to kg/m$^2$ over the length of the sample. In addition to the heat flux from the radiant panel, there is a small flame front approximately 10 to 30~mm in height that serves as the pilot along the length of the same and provides additional energy to the fuel immediately ahead of the flame.  The impact is to a very limited area ahead of the flame, estimated to be less than 5~mm, since the flame is drawn back to the leading edge of the sample (burned portion) by the draft of the radiant panel.  This effect is shown in Figure~\ref{FRP_flamefront}. In Figure~\ref{FRP_flamefront} several thermocouples can be seen. The exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter, were installed by pulling the positive and negative wires through holes drilled through the gypsum wallboard sample and positioning the bead of the thermocouple in a slit in the paper face made with a razor blade.  This was the same installation method that was used in the instrumented gypsum board wall experiments and to measure the ignition temperature of the cone calorimeter samples.  The thermocouples that were installed in the center of the sample at 100~mm intervals beginning at 100~mm from the edge of the sample closest to the radiant panel.  Three replicate experiments were conducted.  The average time that the flame front reached the thermocouples and the temperature of the thermocouple just prior to the painted paper surface of the gypsum wallboard sample igniting is presented in Table~\ref{tab:FRP_timetemp}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Postion & Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (mm) & (kg/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 100 	& 21	& Yes 	& 133	& 315 	 \\
		\hline 200	& 18 	& Yes 	& 198 	& 296 	 \\
		\hline 300	& 14	& Yes 	& 302 	& 279 	 \\
		\hline 400	& 10	& Yes 	& 502 	& 242 	 \\
		\hline 500	&  7	& No 	& NA 	&     	 \\
		\hline
	\end{tabular}
	\caption[Flooring radiant panel experiment flame spread results]{Flooring radiant panel experiment flame spread results including time to ignition at 100~mm intervals and surface temperature just prior to ignition.  The values are averages from three replicate experiments.}
	\label{tab:FRP_timetemp}
\end{table}

Exposing the painted gypsum wallboard samples to the vertical radiant panel provided a wide range physical damage to samples as shown in Figure~\ref{RPGB_SET}.  However there based on visual observations there were no flaming ignitions.  One of the samples at 15~kW/m$^2$, small sections of the paint/paper fuel would peel or partially separate from the gypsum core and would begin to glow.  This process became more prevalent as the heat fluxes increased. There are no photos of the samples that were exposed to 25~kW/m$^2$.  The radiant exposure caused the samples to collapse as they were being removed from the holder.  Figure~\ref{RP_GB_Front} is a graph of the averaged time vs temperature curves for the three replicates at the 7 different heat flux levels.  Notice that the samples exposure at 17.5~kW/m$^2$ and higher have a temperature increase to a peak temperature followed by a decrease.  This trend may be may indicative of combustion of the paint and paper adding to the energy at on the face of the sample.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kg/m$^2$) & (kg/m$^2$)  \\
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time heat release rate per unit area results]{Painted gypsum wallboard heat release rate per unit area results from the cone calorimeter}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

@@

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 9~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and sample nearing the flameout position are provided in Figure~\ref{FRP_Test_Photos}.

@@

@@

\section{Wall Construction Type Experiments}

@@

\section{Wall Position Experiments}
\subsection{Experimental Arrangement}

>>>>

The results from the three different bench scale experiments are presented and discussed in this section.  All three test methods provided some insight into the ignition and the critical heat flux required for ignition or flame spread of the painted paper surface on the gypsum wall board.  The cone calorimeter also provided values for heat release rate.

@@

Three replicate time to ignition experiments were conducted with the cone calorimeter apparatus at each heat flux.  The results of these experiments are shown in Table~\ref{tab:Gypsum_wallboard_timetoign}.  The time to ignition values in the table are the average of the three times.  Note that no ignition occurred at 25~kW/m$^2$.  The exposure at 25~kW/m$^2$ resulted in charring and cracking of the painted paper surface of the gypsum wallboard sample.  This began to occur, on average, at 270 s of exposure time. The samples were exposed to 25~kW/m$^2$ for 1800~s each and no ignition occurred.  In the experiments with heat fluxes of 35~kW/m$^2$ or higher, ignition did occur and the painted paper surface of the gypsum board was consumed, leaving the gypsum board exposed.  There were no surprises in the data with regard to the heat flux exposures.  The time to ignition was reduced as the incident heat flux was increased.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition &  Fire Effect  \\
		\hline (kW/m$^2$) &  & (s)  &   \\
		\hline 25	& No 	& NA 		& Paint \& paper charred \& cracked \\
		\hline 35 	& Yes 	& 73	 	& Paint \& paper consumed \\
		\hline 50	& Yes 	& 35 	 	& Paint \& paper consumed   \\
		\hline 70	& Yes 	& 21 	 	& Paint \& paper consumed \\
		\hline 75	& Yes 	& 17 		& Paint \& paper consumed \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time to ignition results]{Painted gypsum wallboard time to ignition results from the cone calorimeter}
	\label{tab:Gypsum_wallboard_timetoign}
\end{table}

Additional cone calorimeter experiments were conducted in the University of Canterbury Fire Laboratory.  In these experiments the purpose was to obtain an estimate of the temperature of the painted paper surface at the time of ignition.   Samples of painted gypsum wallboard were shipped from the NIST Fire Research Laboratory in Gaithersburg, MD., USA.  In these experiments an additional incident heat flux level of 30~kg/m$^2$ was added to provide another test between the 25 and 35~kW/m$^2$ levels.  The averaged ignition time and temperature pairs are shown in Table~\ref{tab:Gypsum_wallboard_igntemp}.  The average time to ignition values between the two different calorimeters are within $\pm~50~\%$ and $\pm~10~\%$ at incident heat flux values of 35 and 50~kg/m$^2$, respectively.  The average ignition temperatures from the nine experiments ranged from 335 to 375~$^{\circ}$C.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (kW/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 30 	& Yes 	& 112	& 375 	 \\
		\hline 35	& Yes 	& 50 	& 335 	 \\
		\hline 50	& Yes 	& 38 	& 371 	 \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard ignition temperature results]{Painted gypsum wallboard ignition temperature results from exposure in the cone calorimeter}
	\label{tab:Gypsum_wallboard_igntemp}
\end{table}

In the radiant flooring panel experiments the gypsum wallboard was exposed to an incident heat flux that ranged from approximately 21~kW/m$^2$ to 2~kW/m$^2$ over the length of the sample. In addition to the heat flux from the radiant panel, there is a small flame front approximately 10 to 30~mm in height that serves as the pilot along the length of the sample and provides additional energy to the fuel immediately ahead of the flame.  The impact is limited to a small distance ahead of the flame, estimated to be less than 5~mm, since the flame is drawn back to the leading edge of the sample (burned portion) by the draft of the radiant panel.  This effect is shown in Figure~\ref{FRP_flamefront}. In Figure~\ref{FRP_flamefront} several thermocouples can be seen. The exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter, were installed by pulling the positive and negative wires through holes drilled through the gypsum wallboard sample and positioning the bead of the thermocouple in a slit in the painted paper face made with a razor blade.  This was the same installation method that was used in the instrumented gypsum board wall experiments and in the cone calorimeter samples to measure ignition temperature.  The thermocouples were installed in the center of the sample at 100~mm intervals beginning at 100~mm from the edge of the sample closest to the radiant panel.  Three replicate experiments were conducted.  The average time that the flame front reached the thermocouples and the temperature of the thermocouple just prior to the painted paper surface of the gypsum wallboard sample igniting is presented in Table~\ref{tab:FRP_timetemp}.

@@

 With the flooring radiant panel test, the incident heat flux from the panel and the pre-heated gypsum wallboard sample enables flame spread for approximately 500~s.  The temperatures prior to ignition ranged from an average of 315 $^{\circ}$C for the 100~mm position to 242 $^{\circ}$C for the 400~mm position.  The average

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Postion & Incident Heat Flux & Ignition & Time to Ignition & Ignition Temperature \\
		\hline (mm) & (kW/m$^2$) & (Yes or No) & (s)  & Temperature ($^{\circ}$C)   \\
		\hline 100 	& 21	& Yes 	& 133	& 315 	 \\
		\hline 200	& 18 	& Yes 	& 198 	& 296 	 \\
		\hline 300	& 14	& Yes 	& 302 	& 279 	 \\
		\hline 400	& 10	& Yes 	& 502 	& 242 	 \\
		\hline 500	&  7	& No 	& NA 	& NA   	 \\
		\hline
	\end{tabular}
	\caption[Flooring radiant panel experiment flame spread results]{Flooring radiant panel experiment flame spread results including time to ignition at 100~mm intervals and surface temperature just prior to ignition.  The values are averages from three replicate experiments.}
	\label{tab:FRP_timetemp}
\end{table}

Exposing the painted gypsum wallboard samples to the vertical radiant panel provided a wide range physical damage to samples as shown in Figure~\ref{RPGB_SET}.  However based on visual observations there were no flaming ignitions.  For one of the samples at 15~kW/m$^2$, small sections of the painted paper surface on the gypsum wall board would lift or partially separate from the gypsum core and would begin to glow.  This process became more prevalent as the heat fluxes increased.  The radiant exposure of 25~kW/m$^2$ caused the samples to collapse as they were being removed from the holder.  Hence there were no photos of those samples.   Figure~\ref{RP_GB_Front} is a graph of the averaged time vs temperature curves for the three replicates at the 7 different heat flux levels.  Notice that the samples exposed to 17.5~kW/m$^2$ and higher have a temperature increase to a peak temperature followed by a decrease.  This trend may be indicative of the combustion of the paint and paper adding to the energy at the face of the sample. The temperatures at which the change in rate of temperature rise occur are between 300 and 400~ $^{\circ}$C.

@@

The cone calorimeter experiments generated heat release are per unit area values ranging from a low of approximately 160 kW/m$^2$ at an incident heat flux of 35 kW/m$^2$ to approximately 230 kW/m$^2$ at an incident heat flux of 75 kW/m$^2$.  The average total heat released during each burn was 2.8 MJ/m$^2$.

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time heat release rate per unit area results]{Painted gypsum wallboard heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

@@

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 8.9~kW/m$^2$ with a 95$\%$ confidence limit of $\pm$~0.6~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and sample nearing the flameout position are provided in Figure~\ref{FRP_Test_Photos}.

@@

While the vertical radiant panel did not provide ignition data or heat release rate values it did provide information on the fire effect development on the painted gypsum wallboard samples with respect to heat flux.  Referring again to the five fire effect categories listed by DeHann~\cite{DeHaan:2012} and examining the Figure~\ref{RPGB_SET} we see surface thermal effects of discoloration on the painted paper surface of the gypsum wallboard samples at both the 10 and 12.5~kW/m$^2$ incident heat flux exposures.  The samples exposed to 15~kW/m$^2$ exhibit charring and given the thin nature of the painted paper surface penetration was evidenced on two of the three sample.  The sample of the right also shows signs of the painted paper being consumed.  The samples exposed to 17.5~kW/m$^2$ and above all show signs of significant levels of painted paper consumption.  As the heat flux was increased, the area of damage also increased. Even though the incident heat flux values were lower than the characteristic value measured at the center of the sample, as the heat flux in the center increased heat flux at the edges also increased above 15~kW/m$^2$. Hence the damaged area increased outward toward the edges. It would appear in this test series, that 15~kW/m$^2$ is right on the threshold of painted paper penetration and consumption.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

Each apparatus provided similar but different thermal exposure conditions and each test method has a specific parameter or parameters that they were designed to measure for. Currently there is no specific protocol for a fire model user to follow to develop input data.  Nor is there a "standard" catalog of fire model inputs for a given material.  As a result the literature is sourced for information and users may to conduct experiments, such as those above, in order to develop values or ranges of values which could be used in the calculations in Chapters 8 and 9 or to evaluate the validity of the calculations in those chapters.

@@

\section{Wall Position Experiments}

@@
